Modeling electrostatics in molecular biology: A tutorial of DelPhi and associated resources [Article v\versionnumber]

### 2019-06-25 article-manuscript/article-manuscript-tutorial-v01.tex
Remove duplicates

\author[1]{Shailesh Kumar Panday}
\author[1]{Mihiri H.B. Shashikala}
\author[1]{Mahesh Koirala}
\author[1]{Swagata Pahari}
\author[1]{Arghya Chakrvorty}
\author[2]{Yunhui Peng}
\author[3]{Lin Li}
\author[4]{Zhe Jia}
\author[5]{Chuan Li}
\author[1*]{Emil Alexov}
\affil[1]{Department of Physics, Clemson University, Clemson, SC 29634, USA}
\affil[2]{Institution 2}
\affil[3]{Institution 3}
\affil[4]{Institution 4}
\affil[5]{Institution 5}

@@

\orcid{Author 1 name}{AAAA-BBBB-CCCC-DDDD}
\orcid{Author 2 name}{EEEE-FFFF-GGGG-HHHH}

\contrib[\authfn{1}]{These authors contributed equally to this work}
\contrib[\authfn{2}]{These authors also contributed equally to this work}

\presentadd[\authfn{3}]{Department, Institute, Country}
\presentadd[\authfn{4}]{Department, Institute, Country}

@@

Electrostatics pay indispensable role in practically any process in molecular biology. Indeed, at distances larger than several Angstroms all other forces are negligible small and electrostatic force dominates. However, modeling electrostatics in molecular biology is a complicated task due to presence of water phase, mobile ions and irregularly shaped inhomogeneous biological macromolecules. A particular approach to calculating electrostatics in such systems is to apply Poisson-Boltzmann equation (PBE). Here we provide a tutorial for the popular DelPhi package that solves PBE using finite-difference method and delivers the electrostatic potential distribution throughout the modeling box. The tutorial comes with detailed description of different tasks that DelPhi can handle, an assessment of the accuracy against cases with analytical solution and recommendations about DelPhi usage. Furthermore, since electrostatics is a key component of virtually any modeling in molecular biology, we have created many additional resources utilizing DelPhi to model various biology relevant quantities. Tutorials for these resources are also provided along with examples of their usage. 
\end{abstract}

@@

\section{Introduction}
Poisson-Boltzmann equation (PBE) is widely used to describe the electrostatic potential distribution in systems made of biological macromolecules in water phase in presence of mobile ions\cite{xiao2017continuum,li2013progress,jurrus2018improvements,nguyen2017accurate}. Since in this approach the macromolecules are considered using their atomistic 3D structures, it represents an excellent trade off between the speed and the necessary details for accurate description of the relevant phenomena. The traditional implementation of PBE in molecular biology considers that biological macromolecules are low dielectric cavities immersed in water phase which is described as continuum medium with large dielectric constant. Ions are modeled in such a protocol as point charges that obey Boltzmann distribution and can be present outside of so termed Stern Layer (Fig. 1a). Such an implementation of PBE will be termed traditional PBE. In parallel, we have implemented Gaussian-based smooth dielectric function in DelPhi that treats the solute and solvent on the same footage and the entire computational space is described via continuous dielectric function\cite{chakravorty2018gaussian,jia2017treating,li2014modeling,li2013dielectric}. Thus, the macromolecules are considered to be inhomogeneous objects and there is no sharp boundary between solute-solvent. Ions in such a scenario are considered as point charges obeying Boltzmann distribution, but the argument of the Boltzmann function has a desolvation penalty which does not allow ions to propagate into macromolecular interior unless there is a cavity (Fig. 1b). This approach will be termed Gaussian PBE. Both approaches have need used in various computational algorithms aimed at computing biologically important quantities and these resources will be described here as well along with appropriate tutorials.

@@

\subsubsection{Continuum electrostatics quantities} The key electrostatic quantity discussed in this tutorial is electrostatic potential distribution throughout the space. The space is occupied by macromolecule(s) and water phase. In terms of continuum electrostatics, the macromolecules and water phase are considered as continuum media. For such a system, the electrostatic potential obeys various differential equations, which will be described below. Essential term in these equations is so termed dielectric “constant” or dielectric function. In some models the dielectric function is a constant in macromolecular and water media, while in other models it is a smooth continuous function. The macromolecules are considered at atomistic level of details, and atomistic radii and charges are assigned according appropriate force field parameters. Finally, the mobile ions presented in the water phase are considered point charges and their distribution is governed by the Boltzmann law. Below we outline two major models of partitioning the computational space.

\subsubsection{Space partitioning in case of standard two-dielectric model\cite{rocchia2002rapid,baker2001electrostatics}}. Figure 1a shows the space partitioning and the corresponding differential equations being solved in case of traditional PBE. The space is split into three regions labeled as $ \Omega_1 $, $ \Omega_2 $ and $ \Omega_3 $ in Figure \ref{fig:Figure_1}a. The first space region, the $\Omega_1$ is occupied by the macromolecule(s) and its boundary is defined as the molecular surface. It is considered to be a low dielectric cavity with dielectric constant of $\epsilon_1$. Molecular surface itself may be constructed via various approaches, which are not subject of this manuscript. The outer space region, the $\Omega_3$ region, is taken by the bulk water, considered to be continuum medium with dielectric constant of $\epsilon_3$. In-between is space region $\Omega_2$, which is termed Stern layer. This region has the same dielectric properties as the bulk water, but ions are not allowed to penetrate there (Figure \ref{fig:Figure_1}a).

@@

The first region or  is occupied by the macromolecule(s) and its boundary is defined as the molecular surface. This region is considered as a low dielectric cavity of dielectric constant with permanent charges that belong to the macromolecular atoms. Since it does not contain solvent or mobile ions, its potential distribution ($\phi(\Vec{r})$) is governed purely by Poisson equation, as shown:
$$ 
\nabla . \left[ \epsilon(\Vec{r})\nabla\phi(\Vec{r}) \right] = -4\pi\rho(\Vec{r}) 
$$
which can also be rewritten as:
$$ 
\nabla^2\phi(\Vec{r}) = \frac{-4\pi\rho(\Vec{r})}{\epsilon_1} 
$$
where $ \rho(\Vec{r}) $ denotes the charge density of the fixed charges of the macromolecule(s) in question and $ \epsilon_1 $ is the dielectric constant of macromolecule (typically referred as internal dielectric constant).

The outer space region or the $ \Omega_3 $ is occupied by the bulk of the solvent (typically water) which represents a continuum medium with a dielectric value $ \epsilon_3 $. The solvent continuum can also contain mobile ions whose effects are implicitly taken into account using the Boltzmann law. The potential distribution in this regions is governed by the Poisson-Boltzmann equation (PBE). Therefore, here, the charge density, $ \rho_3(\Vec{r}) $, due to the mobile ions (with concentration M) generates a potential distribution that can be expressed as following:
\begin{equation}\label{eqn:pbe_1}
\nabla^2\phi(\Vec{r}) = \frac{-4\pi\rho_3(\Vec{r})}{\epsilon_3} 
\end{equation}
The charge density can be rewritten as:
\begin{align}\label{eqn:pbe_2}
\rho_3(\Vec{r}) &= M_+ e_c - M_- e_c \nonumber \\
&= e_c M \left(exp\left[ - \frac{e_c\phi(\Vec{r})}{k_BT}\right] - exp\left[  \frac{e_c\phi(\Vec{r})}{k_BT}\right]\right) \nonumber \\
&= e_c M \sinh\left({\frac{e_c\phi(\Vec{r})}{k_BT}}\right)
\end{align}
In these equations, the mobile ion charge density is composed of positive and negative charges (from a monovalent salt) whose corresponding concentrations are $ M_+ $ and $ M_- $. Combining the reformatted mobile ion charge density with PBE, the following can be obtained:

@@

In the above equations, $ e_c $ is the elementary charge, $ k_B $ is the Boltzmann constant and $ T $ is the temperature.
In between these two regions exists the ion-exclusion layer (shown as $ \Omega_2 $) which the mobile ions are not allowed to access. Due to the lack of mobile ions as well as the absence of permanent solute charges, this region is governed by Laplace’s equation.
\begin{equation}
\nabla^2\phi(\Vec{r}) = 0
\end{equation}
The ion-exclusion layer or the Stern layer has the same dielectric properties as the solvent region or $ \Omega_3 $ (i.e. $ \epsilon_3 $). Therefore, $ \epsilon_2=\epsilon_3 $ . This indicates the role of the molecular surface as the surface that separates two dielectric media. The molecular surface can be constructed using various approaches but they are out of the scope of this article\cite{rocchia2002rapid,chakravorty2019grid,decherchi2013between}.

\subsubsection{Space partitioning with the Gaussian-based smooth dielectric model}
The Gaussian-based smooth dielectric model is designed to deliver a continuous distribution of dielectric values in space as opposed to representing it as a union of two media with distinct dielectric constants. 
The smooth distribution reflects the effect of restrictions on the mobility of the macromolecular and solvent regions on their polarizability or the dielectric value. The distribution is assigned based on the macromolecular atomic packing density in a given region. Its formulation ensures that the dielectric value is the lowest ($ \epsilon_1 $) at the center of the macromolecular atoms and is the highest ( $ \epsilon_2 $ ) in the bulk. Through this assignment, regions close to the macromolecular atom centers attain a lower dielectric value and those farther way from the macromolecule attain a higher value. Overall, buried or highly densely packed regions of the system feature a lower dielectric value than those which are less packed (details are provided in refs\cite{chakravorty2019grid,decherchi2013between}). This is illustrated in Figure \ref{fig:Figure_1}b.

The key features of Gaussian-based smooth dielectric functions are that the strict surface (like the molecular surface shown in Figure 1a) is replaced by a smooth transition of the dielectric values from the macromolecular interior ($ \Omega_1 $ in Figure 1) to the bulk ($ \Omega_3 $ in Figure 1). The transient values of the dielectric at the solute-solvent interface collectively reflects the increased polarizability of solvent exposed atoms of the macromolecule (compared to those in the core regions) and decreased polarizability of the solvent in their vicinity (compared to those in the bulk). This occurs by virtue of the solute-solvent interactions that play a critical role in balancing the effects of (de)solvation and intramolecular interactions of the solvent exposed macromolecular atoms. 
Due to the continuous dielectric distribution, a clear distinction of macromolecular and solvent regions is no longer necessary and therefore, the same differential equation (PE or PBE) can be used homogeneously. However, the lack of a dielectric boundary invites a modification in the way by which the screening effect of mobile ions is taken into account. In the traditional 2-dielectric approach, the presence of a mobile ion at a point in space (depending on its accessibility) is guided by the electrostatic potential at that point which is expressed using a Boltzmann distribution function (Eqn. \ref{eqn:pbe_2}). In the Gaussian-based approach, an additional penalty term is added to its potential energy at any point in space (for more details see Ref\cite{jia2017treating}). The penalty term is based on the Born-formalism of the energy of transfer of a centrosymmetric ion across two dielectric media. In the case of the Gaussian-model, the penalty term, $ \Delta G_{penalty} $, is the energy of transfer of an ion of valence $ z $ and Van der Waals radius $ R $ from the solvent (with dielectric $ \epsilon(\Vec{r})$) to a point  (with dielectric ) and is expressed in SI units as:

@@

In Figure \ref{fig:plot_saltation_penalty}, the penalty term as a function of position is shown for a solute with a very simple geometry (one sphere of 2Å radius). As one approaches the vicinity of the solute, the penalty increases dramatically, ensuring that the mobile ions cannot penetrate in the regions dominated by solute’s dielectric values. Farther out in the bulk, the penalty term diminishes to 0, thus retrieving the properties of the 2-dielecric approach.

@@

\subsubsection{Finite difference algorithm}
The Poisson-Boltzmann equation is a non-linear $2^{nd}$ order differential equation that is typically solved using numerical schemes since analytical solutions are only possible for cases with very simple geometry. Inhomogeneity of charge distribution and complex geometrical features are prerogative of biomolecules, therefore, different methods of numerically solving the PBE for biomolecular systems have been developed \cite{klapper1986focusing,davis1989solving,cortis1997numerical,baker2001electrostatics,shestakov2002solution,miertuvs1981electrostatic,totrov2001rapid,lu2009adaptive}. 
Delphi solves the PBE using the finite difference (FD) algorithm. The approach rests on discretization of the entire space, $ \Omega=(\Omega_1 \cup \Omega_2 \cup \Omega_3)$, into a set of adjacent cubes, all of the same side length. Along each of the three Cartesian orthogonal directions, identical number of these cubes are placed and their union forms a larger cube which is also referred to as the computational box. Points where two or more cubes meet are known as the grid points and they are the locations where the electrostatic potentials and electric fields are determined.  Between two adjacently placed grid-points, a mid-point is located where the dielectric values are assigned. Collectively, the grid-points and mid-points are all placed within the bounds of the computational box. Figure 3 provides an illustration of this discretization and a computational box with ‘$ M $’ cubes per side, each with a side length of ‘$ h $’.

\begin{figure*}
\includegraphics[width=0.95\linewidth]{Figure_3.png}
\caption{Discretization of the space containing the solvated biomolecule. For the finite-difference approach of solving the PBE, Delphi divides the total space by adjacently placing equal numbers of cubes, each of side length ‘h’, along all the three Cartesian orthogonal directions as shown in \textbf{(a)}. The vertices of a cube are known as the grid-points and are shown in \textbf{(b)} using solid black points. The points in between two adjacent grid-points are called the mid-points where the values of dielectric constants are assigned. They are shown using grey points in \textbf{(c)}.}
\label{fig:grid-scheme-representation}
%% If the optional argument in the square brackets is ``none'', then the caption *will not appear in the main figure at all* and only the full caption will appear under the supplementary figure at the end of the manuscript.
\end{figure*}

@@

By performing a finite volume integration of the above linear PBE and projecting the results in the discretized space, one can derive a numerical formula that establishes a relation of the electrostatic potential, $ \phi_0 $, on a grid-point with that of the six of its neighboring grid-points (see Fig 3(c)). The value also depends on the charge ($ q_0 $) and the modified Debye-Hückel parameter ($ \Bar{\kappa}^2 $) on the grid-point and also on the dielectric values at the six midpoints surrounding it. The numerical formula has the following form:
\begin{equation}
    \phi_0 = 
    \left[ 
        \frac{\sum_i^6{\epsilon_i\phi_i } + 
                \frac{4\pi q_0}{h}
             }
             {\sum_i^6{\epsilon_i} + \Bar{\kappa}^2h^2}
    \right]
\end{equation}
When the Gaussian-based smooth dielectric model is invoked, the FD formula remains the same expression except that it incorporates a space dependent penalty term (equation ???) imposed on the mobile ions. The modified expression has the following form:

@@

\begin{equation}
    G_{grid} = 0.5 \left[ \sum_i^{\text{number of charges}}{\frac{q_i q_j}{\epsilon_{in}r_{ij}}} \right]
\end{equation}	
where $ q_i $ are the real charges, $r_{ij} $ is the distance between charge ‘$i$’ and ‘$j$’ and $ \epsilon_{in} $ is the internal dielectric constant.

@@

\begin{equation}
    G_{rxn} = 0.5 \left[ \sum_i^{\text{number of charges/induced charges}}{\frac{q_i \sigma_j}{\epsilon_{out}r_{ij}}} \right]
\end{equation}
where $ \sigma_j $ are induced surface charges.

The third term represents the energy of interactions between permanent charges and mobile ions. Formally it can be described by formula similar to eq. (3) where instead of induced charges one takes excess ion density. However, this approach was abandoned in the newest DelPhi version because it is very slow and requires small “percent fill”. Instead, we recommend using grid energy difference between runs with salt and without salt. 
\begin{equation}\label{eqn:electro_osmostic_press}
    \Delta \Pi = \sum_i{c_i(e^{-\frac{z_ie\phi(\Vec{r})}{k_BT}} - 1)}
\end{equation}

Finally, if one invokes non-linear PBE, two additional energy terms should be added to the total electrostatic energy. Following the work of Sharp and Honig\cite{sharp1990electrostatic}, they are termed “osmotic pressure” (see Eqn. \ref{eqn:electro_osmostic_press}) and electrostatic "stress”

@@

\newcommand{\angstrom}{\text{\normalfont\AA}}

\begin{figure}[hbt!]
\includegraphics[width=\linewidth]{Figure_4.png}
\caption{The schematic representation of relationship between "percent fill" (\ttvar{perfil}), "grid size" (\ttvar{gsize}), and \ttvar{scale}. The parameter \ttvar{gsize} decides number of grid lines in each direction e.g. \ttvar{gsize} is 9 in given cartoon. The parameter \ttvar{scale} decides number of grid lines per \SI{\angstrom}. If we assume perfil $ pf\% $ and maximum dimension in X-, Y- directions be $ L_{xp} $ \SI{\angstrom} and $ L_{yp} $ \SI{\angstrom} as shown in cartoon, then the size of the grid box $ L_x $ in \SI{\angstrom} is computed using relation $ L_x = \frac{max(L_{xp}, L_{yp}) \times 100}{pf} $.The value of \ttvar{scale} in this case would be $ \ttvar{scale} = \frac{L_x}{\ttvar{gsize}} = \frac{L_x}{9} $.}
\label{fig:perfil-gsize-scale}
\end{figure}

Note that these three parameters, “scale”, “grid size” and “percent fill” are interdependent (Figure \ref{fig:perfil-gsize-scale}). Only two are needed. Users are advised to use “scale” and “percent fill” (\ttvar{perfil}) and to let DelPhi decide the “grid size” (\ttvar{gsize}). In the traditional two-dielectric model, the computational box is split into low dielectric cavities occupied by macromolecule(s) and water phase and appropriate dielectric constant values are assigned to the mid-point belonging to the corresponding phases. In the Gaussian model, coordinates and radii are sued to construct a smooth Gaussian-based dielectric function for the entire computational box without distinguishing be macromolecule(s) from the water phase (for mode details see Refs \cite{li2014modeling,li2013dielectric}). Once this is completed, DelPhi initiates FD iterations to compute electrostatic potential distribution and to calculate the corresponding energy, as it will be outlined below. Several practical advice about DelPhi parameters are given here:

@@

    \item Linear versus non-linear PBE. DelPhi allows the user to select what form of PBE to be solved: linear or non-linear. Solving non-linear PBE is typically more time consuming and should be avoided if not necessary. Typically, if the system does not have large net charge, there is no need to invoke non-linearity\cite{talley2008electrostatic,chakavorty2016electrostatic}. Another way to assess the importance of non-linearity is to
\end{enumerate}

@@

The other example of how to calculate polar solvation energy of a protein, all the related files are kept in directory  and \ttvar{Example_3.1.1/Ex2/}. The DelPhi run for computing it via traditional 2-dielectric approach can be initiated using command below:

>>>>

\author[1]{Shailesh Kumar Panday}
\author[1]{Mihiri H.B. Shashikala}
\author[1]{Mahesh Koirala}
\author[1]{Swagata Pahari}
\author[1]{Arghya Chakrvorty}
\author[1]{Yunhui Peng}
\author[2]{Lin Li}
\author[1]{Zhe Jia}
\author[3]{Chuan Li}
\author[1,*]{Emil Alexov}
\affil[1]{Department of Physics and Astronomy, Clemson University, Clemson, SC 29634}
\affil[2]{Department of Physics, University of Texas at EI Paso, TX 79968}
\affil[3]{Department of Mathematics, West Chester University of Pennsylvania, West Chester, PA 19383}

@@

@@

Electrostatics play indispensable role in practically any process in molecular biology. Indeed, at distances larger than several Angstroms all other forces are negligibly small and electrostatic force dominates. However, modeling electrostatics in molecular biology is a complicated task due to presence of water phase, mobile ions and irregularly shaped inhomogeneous biological macromolecules. A particular approach to calculating electrostatics in such systems is to apply Poisson-Boltzmann equation (PBE). Here we provide a tutorial for the popular DelPhi package that solves PBE using finite-difference method and delivers the electrostatic potential distribution throughout the modeling box. The tutorial comes with detailed description of different tasks that DelPhi can handle, an assessment of the accuracy against cases with analytical solution and recommendations about DelPhi usage. Furthermore, since electrostatics is a key component of virtually any modeling in molecular biology, we have created many additional resources utilizing DelPhi to model various biology relevant quantities. Tutorials for these resources are also provided along with examples of their usage. 
\end{abstract}

@@

\section{Introduction}
Poisson-Boltzmann equation (PBE) is widely used to describe the electrostatic potential distribution in systems made of biological macromolecules in water phase in presence of mobile ions\cite{xiao2017continuum,li2013progress,jurrus2018improvements,nguyen2017accurate}. Since in this approach the macromolecules are considered using their atomistic 3D structures, it represents an excellent trade off between the speed and the necessary details for accurate description of the relevant phenomena. The traditional implementation of PBE in molecular biology considers that biological macromolecules are low dielectric cavities immersed in water phase which is described as continuum medium with large dielectric constant. Ions are modeled in such a protocol as point charges that obey Boltzmann distribution and can be present outside of so termed Stern Layer (Figure \ref{fig:Figure_1}a). Such an implementation of PBE will be termed traditional PBE. In parallel, we have implemented Gaussian-based smooth dielectric function in DelPhi that treats the solute and solvent on the same footage and the entire computational space is described via continuous dielectric function\cite{chakravorty2018gaussian,jia2017treating,li2014modeling,li2013dielectric}. Thus, the macromolecules are considered to be inhomogeneous objects and there is no sharp boundary between solute-solvent. Ions in such a scenario are considered as point charges obeying Boltzmann distribution, but the argument of the Boltzmann function has a desolvation penalty which does not allow ions to propagate into macromolecular interior unless there is a cavity (Figure \ref{fig:Figure_1}b). This approach will be termed Gaussian PBE. Both approaches have been used in various computational algorithms aimed at computing biologically important quantities and these resources will be described here as well along with appropriate tutorials.

@@

\subsubsection{Continuum electrostatics quantities} The key electrostatic quantity discussed in this tutorial is electrostatic potential distribution throughout the space. The space is occupied by macromolecule(s) and water phase. In terms of continuum electrostatics, the macromolecules and water phase are considered as continuum media. For such a system, the electrostatic potential obeys various differential equations, which will be described below. Essential term in these equations is so termed dielectric “constant” or dielectric function. In some models the dielectric function is a constant in macromolecular and water media, while in other models it is a smooth continuous function. The macromolecules are considered at atomistic level of details, and atomistic radii and charges are assigned according appropriate force field parameters. Finally, the mobile ions present in the water phase are considered point charges and their distribution is governed by the Boltzmann law. Below we outline two major models of partitioning the computational space.

\subsubsection{Space partitioning in case of standard two-dielectric model\cite{rocchia2002rapid,baker2001electrostatics}}. The space partitioning and the corresponding differential equations being solved in case of traditional PBE involve splitting the space into three regions labeled as $ \Omega_1 $, $ \Omega_2 $ and $ \Omega_3 $ in Figure \ref{fig:Figure_1}a. The first space region, the $\Omega_1$ is occupied by the macromolecule(s) and its boundary is defined as the molecular surface. It is considered to be a low dielectric cavity with dielectric constant of $\epsilon_1$. Molecular surface itself may be constructed via various approaches, which are not subject of this manuscript. The outer space region, the $\Omega_3$ region, is taken by the bulk water, considered to be continuum medium with dielectric constant of $\epsilon_3$. In-between is space region $\Omega_2$, which is termed Stern layer. This region has the same dielectric properties as the bulk water, but ions are not allowed to penetrate there (Figure \ref{fig:Figure_1}a).

@@

The first region of space $ \Omega_1 $, is occupied by the macromolecule(s) and its boundary is defined as the molecular surface. This region is considered as a low dielectric cavity of dielectric constant with permanent charges that belong to the macromolecular atoms. Since it does not contain solvent or mobile ions, its potential distribution ($\phi(\Vec{r})$) is governed purely by Poisson equation, as shown:

\begin{equation}
 \nabla . \left[ \epsilon(\Vec{r})\nabla\phi(\Vec{r}) \right] = -4\pi\rho(\Vec{r})    
\end{equation} 
which can also be rewritten as:
\begin{equation}
\nabla^2\phi(\Vec{r}) = \frac{-4\pi\rho(\Vec{r})}{\epsilon_1} 
\end{equation}
where $ \rho(\Vec{r}) $ denotes the charge density of the fixed charges of the macromolecule(s) in question and $ \epsilon_1 $ is the dielectric constant of macromolecule (typically referred as internal dielectric constant).

The outer space region or the $ \Omega_3 $ is occupied by the bulk of the solvent (typically water) which represents a continuum medium with a dielectric value $ \epsilon_3 $. The solvent continuum can also contain mobile ions whose effects are implicitly taken into account using the Boltzmann law. The potential distribution in this regions is governed by the Poisson-Boltzmann equation (PBE). Therefore, here, the charge density, $ \rho_3(\Vec{r}) $, due to the mobile ions (with concentration M) generates a potential distribution that can be expressed as Eqn. \ref{eqn:pbe_1}.

\begin{equation}\label{eqn:pbe_1}
\nabla^2\phi(\Vec{r}) = \frac{-4\pi\rho_3(\Vec{r})}{\epsilon_3} 
\end{equation}

The charge density can be rewritten as Eqn. \ref{eqn:pbe_2}:
\begin{align}\label{eqn:pbe_2}
\rho_3(\Vec{r}) &= M_+ e_c - M_- e_c \nonumber \\
&= e_c M \left(exp\left[ - \frac{e_c\phi(\Vec{r})}{k_BT}\right] - exp\left[  \frac{e_c\phi(\Vec{r})}{k_BT}\right]\right) \nonumber \\
&= e_c M \sinh\left({\frac{e_c\phi(\Vec{r})}{k_BT}}\right)
\end{align}
In these equations, the mobile ion charge density is composed of positive and negative charges (from a monovalent salt) whose corresponding concentrations are $ M_+ $ and $ M_- $. Combining the reformatted mobile ion charge density with PBE, the following can be obtained:

@@

In the above equations, $ e_c $ is the elementary charge, $ k_B $ is the Boltzmann constant and $ T $ is the absolute temperature.
In between these two regions exists the ion-exclusion layer (shown as $ \Omega_2 $) which the mobile ions are not allowed to access. Due to the lack of mobile ions as well as the absence of permanent solute charges, this region is governed by Laplace’s equation.
\begin{equation}
\nabla^2\phi(\Vec{r}) = 0
\end{equation}
The ion-exclusion layer or the Stern layer has the same dielectric properties as the solvent region or $ \Omega_3 $ (i.e. $ \epsilon_3 $). Therefore, $ \epsilon_2=\epsilon_3 $ . This indicates the role of the molecular surface as the surface that separates two dielectric media. The molecular surface can be constructed using various approaches but they are out of the scope of this article\cite{rocchia2002rapid,chakravorty2019grid,decherchi2013between}.

\subsubsection{Space partitioning with the Gaussian-based smooth dielectric model}
The Gaussian-based smooth dielectric model is designed to deliver a continuous distribution of dielectric values in space as opposed to representing it as a union of two media with distinct dielectric constants. 
The smooth distribution reflects the effect of restrictions on the mobility of the macromolecular and solvent regions on their polarizability or the dielectric value. The distribution is assigned based on the macromolecular atomic packing density in a given region. Its formulation ensures that the dielectric value is the lowest ($ \epsilon_1 $) at the center of the macromolecular atoms and is the highest ( $ \epsilon_2 $ ) in the bulk. Through this assignment, regions close to the macromolecular atom centers attain a lower dielectric value and those farther away from the macromolecule attain a higher value. Overall, buried or highly densely packed regions of the system feature a lower dielectric value than those which are less packed (details are provided in refs\cite{chakravorty2019grid,decherchi2013between}). This is illustrated in Figure \ref{fig:Figure_1}b.

The key features of Gaussian-based smooth dielectric functions are that the strict surface (like the molecular surface shown in Figure \ref{fig:Figure_1}a) is replaced by a smooth transition of the dielectric values from the macromolecular interior $ \Omega_1 $ to the bulk $ \Omega_3 $ as shown in Figure \ref{fig:Figure_1}). The transient values of the dielectric at the solute-solvent interface collectively reflects the increased polarizability of solvent exposed atoms of the macromolecule (compared to those in the core regions) and decreased polarizability of the solvent in their vicinity (compared to those in the bulk). This occurs by virtue of the solute-solvent interactions that play a critical role in balancing the effects of (de)solvation and intramolecular interactions of the solvent exposed macromolecular atoms. 
Due to the continuous dielectric distribution, a clear distinction of macromolecular and solvent regions is no longer necessary and therefore, the same differential equation (PE or PBE) can be used homogeneously. However, the lack of a dielectric boundary invites a modification in the way by which the screening effect of mobile ions is taken into account. In the traditional 2-dielectric approach, the presence of a mobile ion at a point in space (depending on its accessibility) is guided by the electrostatic potential at that point which is expressed using a Boltzmann distribution function (Eqn. \ref{eqn:pbe_2}). In the Gaussian-based approach, an additional penalty term is added to its potential energy at any point in space (for more details see Ref\cite{jia2017treating}). The penalty term is based on the Born-formalism of the energy of transfer of a centrosymmetric ion across two dielectric media. In the case of the Gaussian-model, the penalty term, $ \Delta G_{penalty} $, is the energy of transfer of an ion of valence $ z $ and Van der Waals radius $ R $ from the solvent (with dielectric $ \epsilon(\Vec{r})$) to a point  (with dielectric ) and is expressed in SI units as:

@@

In Figure \ref{fig:plot_saltation_penalty}, the penalty term as a function of position is shown for a solute with a very simple geometry (one sphere of 2 \text{\AA} radius). As one approaches the vicinity of the solute, the penalty increases dramatically, ensuring that the mobile ions cannot penetrate in the regions dominated by solute’s dielectric values. Farther out in the bulk, the penalty term diminishes to zero, thus retrieving the properties of the 2-dielecric approach.

@@

\subsubsection{Finite difference algorithm}
The Poisson-Boltzmann equation is a non-linear $2^{nd}$ order differential equation that is typically solved using numerical schemes since analytical solutions are only possible for cases with very simple geometry. Inhomogeneity of charge distribution and complex geometrical features are prerogative of biomolecules, therefore, different methods of numerically solving the PBE for biomolecular systems have been developed \cite{klapper1986focusing,davis1989solving,cortis1997numerical,baker2001electrostatics,shestakov2002solution,miertuvs1981electrostatic,totrov2001rapid,lu2009adaptive}. 
DelPhi solves the PBE using the finite difference (FD) algorithm. The approach rests on discretization of the entire space, $ \Omega=(\Omega_1 \cup \Omega_2 \cup \Omega_3)$, into a set of adjacent cubes, all of the same side length. Along each of the three Cartesian orthogonal directions, identical number of these cubes are placed and their union forms a larger cube which is also referred to as the computational box. Points where two or more cubes meet are known as the grid points and they are the locations where the electrostatic potentials and electric fields are determined.  Between two adjacently placed grid-points, a mid-point is located where the dielectric values are assigned. Collectively, the grid-points and mid-points are all placed within the bounds of the computational box. Figure \ref{fig:grid-scheme-representation}, provides an illustration of this discretization and a computational box with ‘$ M $’ cubes per side, each with a side length of ‘$ h $’.

\begin{figure*}
\includegraphics[width=0.95\linewidth]{Figure_3.png}
\caption{Discretization of the space containing the solvated biomolecule. For the finite-difference approach of solving the PBE, DelPhi divides the total space by adjacently placing equal numbers of cubes, each of side length ‘h’, along all the three Cartesian orthogonal directions as shown in \textbf{(a)}. The vertices of a cube are known as the grid-points and are shown in \textbf{(b)} using solid black points. The points in between two adjacent grid-points are called the mid-points where the values of dielectric constants are assigned. They are shown using grey points in \textbf{(c)}.}
\label{fig:grid-scheme-representation}
%% If the optional argument in the square brackets is ``none'', then the caption *will not appear in the main figure at all* and only the full caption will appear under the supplementary figure at the end of the manuscript.
\end{figure*}

@@

By performing a finite volume integration of the above linear PBE and projecting the results in the discretized space, one can derive a numerical formula that establishes a relation of the electrostatic potential, $ \phi_0 $, on a grid-point with that of the six of its neighboring grid-points (see Figure \ref{fig:grid-scheme-representation}(c)). The value also depends on the charge ($ q_0 $) and the modified Debye-Hückel parameter ($ \Bar{\kappa}^2 $) on the grid-point and also on the dielectric values at the six midpoints surrounding it. The numerical formula has the following form:
\begin{equation}
    \phi_0 = 
    \left[ 
        \frac{\sum_i^6{\epsilon_i\phi_i } + 
                \frac{4\pi q_0}{h}
             }
             {\sum_i^6{\epsilon_i} + \Bar{\kappa}^2h^2}
    \right]
\end{equation}
When the Gaussian-based smooth dielectric model is invoked, the FD formula remains the same expression except that it incorporates a space dependent penalty term imposed on the mobile ions. The modified expression has the following form:

@@

\begin{equation}
    G_{Coul} = 0.5 \left[ \sum_i^{\text{number of charges}}{\frac{q_i q_j}{\epsilon_{in}r_{ij}}} \right]
\end{equation}	
where $ q_i $ are the real charges, $r_{ij} $ is the distance between charge ‘$i$’ and ‘$j$’ and $ \epsilon_{in} $ is the internal dielectric constant.

@@

\begin{equation}\label{eqn:G_rxn}
    G_{rxn} = 0.5 \left[ \sum_{i, j}^{\text{number of charges/induced charges}}{\frac{q_i \sigma_j}{\epsilon_{out}r_{ij}}} \right]
\end{equation}
where $ \sigma_j $ are induced surface charges.

The third term represents the energy of interactions between permanent charges and mobile ions. Formally it could be described by formula similar to Eqn. \ref{eqn:G_rxn}, where instead of induced charges one takes excess ion density. However, this approach was abandoned in the newest DelPhi version because it is very slow and requires small “percent fill”. Instead, we recommend using grid energy difference between runs with salt and without salt.

Finally, if one invokes non-linear PBE, two additional energy terms should be added to the total electrostatic energy. Following the work of Sharp and Honig\cite{sharp1990electrostatic}, they are termed “osmotic pressure” and electrostatic "stress”.

@@

\begin{figure}[hbt!]
\includegraphics[width=\linewidth]{Figure_4.png}
\caption{The schematic representation of relationship between "percent fill" (\ttvar{perfil}), "grid size" (\ttvar{gsize}), and \ttvar{scale}. The parameter \ttvar{gsize} decides number of grid lines in each direction e.g. \ttvar{gsize} is 9 in given cartoon. The parameter \ttvar{scale} decides number of grid lines per \text{\AA}. If we assume perfil $ pf\% $ and maximum dimension in X-, Y- directions be $ L_{xp} $ \text{\AA} and $ L_{yp} $ \text{\AA} as shown in cartoon, then the size of the grid box $ L_x $ in \text{\AA} is computed using relation $ L_x = \frac{max(L_{xp}, L_{yp}) \times 100}{pf} $. The value of \ttvar{scale} in this case would be $ \ttvar{scale} = \frac{L_x}{\ttvar{gsize}} = \frac{L_x}{9} $.}
\label{fig:perfil-gsize-scale}
\end{figure}

Note that these three parameters, “scale”, “grid size” and “percent fill” are interdependent (Figure \ref{fig:perfil-gsize-scale}). Only two are needed. Users are advised to use “scale” and “percent fill” (\ttvar{perfil}) and to let DelPhi decide the “grid size” (\ttvar{gsize}). In the traditional two-dielectric model, the computational box is split into low dielectric cavities occupied by macromolecule(s) and water phase and appropriate dielectric constant values are assigned to the mid-point belonging to the corresponding phases. In the Gaussian model, coordinates and radii are used to construct a smooth Gaussian-based dielectric function for the entire computational box without distinguishing the macromolecule(s) from the water phase (for mode details see Refs \cite{li2014modeling,li2013dielectric}). Once this is completed, DelPhi initiates FD iterations to compute electrostatic potential distribution and to calculate the corresponding energy, as it will be outlined below. Several practical advice about DelPhi parameters are given here:

@@

    \item Linear versus non-linear PBE. DelPhi allows the user to select what form of PBE to be solved: linear or non-linear. Solving non-linear PBE is typically more time consuming and should be avoided if not necessary. Typically, if the system does not have large net charge, there is no need to invoke non-linearity\cite{talley2008electrostatic,chakavorty2016electrostatic}.
\end{enumerate}

@@

The other example of how to calculate polar solvation energy of a protein, all the related files are kept in the directory \ttvar{Example_3.1.1/Ex2/}. The DelPhi run for computing it via traditional 2-dielectric approach can be initiated using command below:
